Fast Sorting on Modern Computers

### 2005-07-26 docs/tech_report/heap.tex
minor changes as rereading

\section{Results}
All simulations presented here were performed on 32 byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 bytes, we stick with 8-byte cache lines. \verb!UINT_MAX! was used
as the sentinel.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches are predictable. An 8-heap has seven
comparisons between children, with the result that only an eighth of the branches
are predictable.

@@

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, with the 8-heap having the fewest misses.
This continues into the level 2 cache, where the flattest slope belongs to the
8-heap. The 4-heap with the 32-byte line has fewer misses than the 16-byte
version, which is surprising since the 16-byte version can fit twice as many
blocks in the cache, and uses each block fully.  This may be due to the extra
compulsory misses due to having twice as many blocks to load. The difference
between fully associative and direct mapped is only minor.

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have approximately half as many level 2 cache misses
as the out-of-place sorts (See Figure \ref{heap level 2 misses}).

The branch prediction behaves similarly to the expected results, in that the
prediction ratio gets worse as the fanout gets larger. However, the difference
is minor between the 4-heap and 8-heap, and the difference for larger fanout
would be less still. The difference between the 2-heap and 4-heap result is
interesting though, the 2-heap has more branches, but less misses per key.

@@

Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).}. Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

>>>>

\section{Results}
All simulations presented here were performed on 32 byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 ints, we stick with 32 byte cache lines. \verb!UINT_MAX! was used
as the sentinel.

@@

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches logn}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

@@

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, with the 8-heap having the fewest misses.
This continues into the level 2 cache, where the flattest slope belongs to the
8-heap. The difference between fully associative and direct mapped is only
minor, with the fully associative performing slightly better than the direct
mapped cache.

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have no compulsary misses\footnote{See Section
\ref{TODO}.}, while the out-of-place sorts have one compulsory miss in eight,
when the auxilliary heap is being populated.

The branch prediction behaves differently to the expected results, as the
prediction ratio gets worse as the fanout gets larger. The reason we predicted
differently was due to selection sort's $O(log N)$ mispredictions; however, this
is over the entire length of the array, and it is not accurate to extrapolate
these results over a sequence only seven keys long.

The difference between the 4-heap and 8-heap is minor , and the difference for
larger fanout would be less still. The difference between the 2-heap and 4-heap
result is interesting though, the 2-heap has more branches, but less misses per
key.

@@

%TODO move this to testing section
Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).}. Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.
